---
title: "Genomics/High Dim Hw1"
author: "Wei-Yu Tseng"
date: "9/4/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Question 1.B:  

```{r}
famuss <- read.csv("https://raw.githubusercontent.com/xuranw/469_public/master/data/famuss.csv")
```
a.    
```{r}
ncol(famuss)
colnames(famuss)[c(1,81:83)]
```
Although *famus* dataset has 83 columns, 4 of them are not SNP. 
There are 79 different SNPs present in the dataset.  
b.  
```{r}
which(colnames(famuss)=='actn3_1671064')
```
The 4th column contains the information about the **actn3_1671064** SNP.  
  
c.
```{r}
table(famuss[,4])
```
There are 3 different minor genotypes, with 169 observations of type **AA**, 262 observations of type **GA**, and 107 observations of **GG** out of a total of 531 observations.  
d.
```{r}
table(famuss[,c(4,83)])
```
The table suggests that most of Caucasians have allele "A".  
  
Question 1.C
a.  
```{r}
table = table(famuss[,'esr1_rs2077647'])
names(table) = c(1,0,2)
barplot(sort(table,decreasing = TRUE),main = 'Genotype counts of SNP esr1_rs2077647', xlab = 'Numeric coding for SNP esr1_rs2077647', ylab = 'Count')

```
  
Q 2.A  
a.
```{r}
source("https://raw.githubusercontent.com/xuranw/469_public/master/hw1/clt.R")
print(generate_data)
```
The description of each line of code in this functions is listed below:  
Line 1.This function won't work if n isn't a positive integer.  
Line 2.Randomly generate n numbers from 1 to 3.  
Line 3.Create a list with n empty slots.   
Line 4.Generate (the number of total counts of 1 in index_vec)  of samples from $N(10,1)$ and store them to the list.  
Line 5.Generate (the number of total counts of 2 in index_vec)  of samples from $\gamma(2,2)$ and store them to the list.  
Line 6.Generate (the number of total counts of 2 in index_vec)  of samples from $\chi^2(2,2)$ and store them to the list. 
Line 7.Return the list.  
  
In general, the function will randomly create n samples from $N(10,1)$, $\gamma(2,2)$ and $\chi^2(2,2)$ and store them to a list.  
Question 2.B  
```{r}
set.seed(100)
CLT = function(n, tries=10000)
{
  data = rep(NA, tries)
for (i in 1: tries){
  data[i]=mean(generate_data(n))
}
  hist(data, breaks = 100, main = paste('Sample size:',n),xlab = 'Value')
}
par(mfrow=c(2,3))
invisible(lapply(c(1,2,3,5,10,100), CLT))
```


